Reinforcement Learning for Options on Target Volatility Funds

### 2021-05-18 tex/targetvol.tex
Paper: done RL section

\documentclass[11pt]{article}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[outer=4cm, marginparwidth=4.5cm]{geometry}
%\usepackage{marginnote}
\usepackage{amssymb,amsmath,amsthm,latexsym,amsfonts,amscd,dsfont,enumerate,bbm,mathabx}
\usepackage{graphicx,tikz}
\usepackage{draftwatermark}
\usepackage{enumerate} 
% \usepackage{listings}
% \usepackage{booktabs}
\usepackage{mathtools}
% \usepackage[gen]{eurosym}
% \usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage{soul}
\usepackage{comment}

@@

The paper is organized ad follows. In \Cref{sec:TVS} we describe the dynamics of a TVS in presence of valuation adjustments. Then, in \Cref{sec:Derivative} we derive the analytical results, and in \Cref{sec:RL} we illustrate how we have applied RL to solve the dynamic control problem. We conclude the paper with \Cref{sec:Numerical_Results} where we present the numerical results obtained in this work.

@@

 \subsection{Episode Description: Agent-Environment Interaction}
 We consider an episode $\tau$ of length $m+1$ that takes place on a discrete time-grid of fixing times $\mathcal{T}:=\{T_0,...,T_k,...,T_m\}$ with $T_0:=0$ and $T_m=T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^\theta_k$.
 In our case the agent can choose the composition of the risky asset portfolio, so that the policy is the asset allocation weights $\alpha$. From the HJB equation we have derived in section \Cref{subsec:HJB}, we select as state of th environment the following block
 \begin{equation}
 	s_{k} \coloneqq \left[T_k,I_{T_k}, S_{T_k}\right] \quad \forall T_k \in \mathcal{T}.
 \label{eq:state}\end{equation}
 This state contains all the informations needed by the agent to take an optimal action, leading to the maximum option price on the TVS. 
 Once the agent has selected the action $a_k$, it receives the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment defined by
 \begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)],
 \label{eq:reward}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is the BS price as proxy of the residual option price defined by
\begin{equation}
	V_{\text{BS}}(T_k) \coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}}),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \quad \text{and} \quad V(T_0)=0
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} calculated in the state $s_k$. The parameter $\gamma$ plays the role of a discount rate in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future. Choosing $\gamma<1$ introduces a bias in the final result.
If we take the cumulative reward per episode and set $\gamma =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} = \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we think that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Policy and Action Parameterization}
As discussed before, PPO parameterizes the agent policy by means of an artificial neural network. In our case we use a feed forward neural network (FFNN), a network with an acyclic topology and completely characterized by the following hyper-parameters: number of hidden layers $N$, number of neurons per hidden layer $M$, activation function per hidden layer and output activation function. The number of output neurons is fixed by the action space dimension, which is the number of assets $n$ in the risky portfolio, while the number of  input neurons is determined by the dimensionality of the state \eqref{eq:state}, $n+2$.

\begin{comment}
\section{Numerical Investigation}
Dire qui che abbiamo lavorato con l'implementazione di OpenAi

>>>>

\documentclass[11pt]{article}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[outer=3.75cm, marginparwidth=4.5cm]{geometry}
%\usepackage{marginnote}
\usepackage{amssymb,amsmath,amsthm,latexsym,amsfonts,amscd,dsfont,enumerate,bbm,mathabx}
\usepackage{graphicx,tikz}
\usepackage{draftwatermark}
\usepackage{enumerate} 
% \usepackage{listings}
% \usepackage{booktabs}
\usepackage{mathtools}
% \usepackage[gen]{eurosym}
% \usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage{soul}
\usepackage{comment}
\usepackage{bm}

@@

The paper is organized ad follows. In \Cref{sec:TVS} we describe the dynamics of a TVS in presence of valuation adjustments. Then, in \Cref{sec:Derivative} we derive the analytical results, and in \Cref{sec:RL} we illustrate how we have applied RL to solve the dynamic control problem. We conclude the paper with \Cref{sec:Numerical_results} where we present the numerical results obtained in this work.

@@

 \subsection{Episode Description: Agent-Environment Interaction}
 We consider an episode $\tau$ of length $m+1$ that takes place on a discrete time-grid of fixing times $\mathcal{T}:=\{T_0,...,T_k,...,T_m\}$ with $T_0:=0$ and $T_m=T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^\theta_k$.
 In our case the agent can choose the composition of the risky asset portfolio, so that the policy is the asset allocation weights $\alpha$. From the HJB equation we have derived in section \Cref{subsec:HJB}, we select as state of th environment the following block
 \begin{equation}
 	s_{k} \coloneqq \left[T_k,I_{T_k}, S_{T_k}\right] \quad \forall T_k \in \mathcal{T}.
 \label{eq:state}\end{equation}
 This state contains all the informations needed by the agent to take an optimal action, leading to the maximum option price on the TVS. 
 Once the agent has selected the action $a_k$, it receives the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment defined by
 \begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)],
 \label{eq:reward}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is the BS price as proxy of the residual option price defined by
\begin{equation}
	V_{\text{BS}}(T_k) \coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}}),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \quad \text{and} \quad V(T_0)=0
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} calculated in the state $s_k$. The parameter $\gamma$ plays the role of a discount rate in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future. Choosing $\gamma<1$ introduces a bias in the final result.
If we take the cumulative reward per episode and set $\gamma =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we think that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Policy and Action Parameterization}
As discussed before, PPO parameterizes the agent policy by means of an artificial neural network. In our case we use a feed forward neural network (FFNN), a network with an acyclic topology and completely characterized by the following hyper-parameters: number of hidden layers $N$, number of neurons per hidden layer $M$, activation function per hidden layer and output activation function. The number of output neurons is fixed by the action space dimension, which is the number of assets $n$ in the risky portfolio, while the number of  input neurons is determined by the dimensionality of the state \eqref{eq:state}, $n+2$. We work with a stochastic policy whose functional form is a multivariate diagonal Gaussian distribution where the mean $\mu^\theta(s_k)$ is the output vector of the FFNN and the log-standard deviation $\log\sigma$ is an external parameter
\begin{equation}
	\pi^\theta(s_k) \sim \mathcal{N}(\mu^\theta(s_k),e^{\log\sigma}).
\end{equation}
In the \textsc{Baselines} implementation of PPO, the parameter $\log\sigma$ is state-independent, but it decreases as the number of PPO update iterations increases. The idea is that the log-standard deviation will be higher at the beginning of the training phase in order to guarantee a good exploration of the action space while it will be lower at the ending phase to avoid too much noise in the proximity of the optimal policy.

Another important topic is how to parameterize the agent action in order to satisfy some specific constraints on the risky asset allocation weights. In fact, if on one hand with free $\alpha$ it is sufficient to set bounds in the action space, on the other hand the presence of constraints on the risky strategy imposes the need to parameterize the output of the ANN.
In this work we study the constraint where the agent can invest in the risky market adopting long positions only:
\begin{equation}
	\alpha \cdot \mathds{1} = 1 \quad \alpha^i >0 \quad \forall i=1,\dots,n.
\end{equation} 
We study two different action parameterizations for this kind of constraint.
\begin{itemize}
	\item \textbf{$\bm{(n-1)}$-spherical angles}: we rewrite $\alpha^j = (x^i)^2$ and we use $(n-1)$-spherical coordinates with unitary radius. Thus the strategy is parameterized by a vector of angles $\phi = (\phi^1,\dots,\phi^{n-1})^T$ as follows
	\begin{equation}
		\begin{aligned}
			\alpha^1 &\coloneqq \cos^2 (\phi^1), \\
			\alpha^2 &\coloneqq \sin^2(\phi^1)\cos^2(\phi^2),\\
			&\vdots \\
			\alpha^{n-1} &\coloneqq \sin^2(\phi^1)\dots\sin^2(\phi^{n-2})\cos^2(\phi^{n-1}),\\
			\alpha^n &\coloneqq \sin^2(\phi^1)\dots\sin^2(\phi^{n-2})\sin^2(\phi^{n-1}).
		\end{aligned}
		\label{eq:euler_angles}\end{equation}
	where $\phi^i\in[0,\pi/2]$ to avoid oscillations in $\alpha$ that could slow down the training. Thus, with this parameterization, the ANN sample a vector of angles ($a = \phi$) and thus the action space is reduced by one dimension.
	\item \textbf{Normalized action}: we restrict the agent actions to a subset of $(\mathbb{R}^+)^n$ and then normalize the action to obtain the corresponding risky allocation strategy through the formula
	\begin{equation}
		\alpha_t = \frac{a_t}{a_t \cdot \mathds{1}}.
		\label{eq:normalized_action}\end{equation}
\end{itemize}

\section{Numerical Investigation}\label{sec:Numerical_results}
In this section As we have seen in 
\subsection{Black and Scholes Dynamics}
\subsection{Local Volatility Dynamics}
\section{Discussion}
\begin{comment}

--scrivere input normalization della rete
--introduzione sui modelli fatti 
--dire perchè black e del fine tuning dei parametri

\ref{sec:target_volatility_option_derivative_pricing}, the bank evaluation of the most conservative call option price strictly depends on the dynamics followed by the equities underlying the \acrshort{tvs}; in particular the volatility term of the securities affects the \acrshort{tvs} price process \eqref{TVS_last}. In our work we will try to solve this pricing problem through the \acrshort{rl} technique, training the agent in two different environments corresponding to two different mathematical models for the dynamics of financial markets: the time-dependent \acrlong{bs} model and the \acrlong{lv} model. Although the first one is quite simple (its parameters depends only on time) and it is not completely able to represent the complex dynamics of real markets, we have seen that, under Black's conditions, we are able to derive the closed form solution for the optimal strategy \eqref{optimal_strategy}. Thus we can use the \acrlong{bs} model as a test to control if we have built correctly the \acrlong{rl} algorithm; in other words we will check if the \acrshort{rl} agent will take decisions in agreement with the closed form. On the other hand, the \acrfull{lv} model depends not only on time but also on the spot prices of the equities and thus a closed formula for the optimal strategy is not available; we will try to recover it by our \acrshort{rl} algorithm.
